---
title: "Credit Risk and Macroeconomic Indicators in Bulgaria 
"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo=FALSE)

```

```{r, echo=FALSE}

## install libraries if missing
list.of.packages <- c('ggplot2','WDI','eurostat','reshape2',
                      'forecast','car','ggfortify','scales','lubridate',
                      'imputeTS')
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

rm(list=ls())

library(WDI)
library(ggplot2)
library(eurostat)
library(reshape2)
library(forecast)
library(car)
library(ggfortify)
library(scales)
library(lubridate)
library(imputeTS)

setwd('C:/Users/Kimon/Documents/Dennis/IFRS9/Statistics')

graph <- function(df,ylab,title) {
  p1 <- ggplot(data=df, aes(x=df[,1], y=df[,2])) +
    scale_x_date(labels = date_format("%Y"), date_breaks='1 year') +
    geom_line(colour="blue", size=0.75) + 
    xlab('Year') + ylab(ylab) +
    theme(axis.text=element_text(size=10,face="bold",angle=90,hjust=1),
          axis.title=element_text(size=12,face="bold"),
          title=element_text(size=14,face="bold"))+
    ggtitle(title) 

  p2 <- ggplot(data=df, aes(x=df[,1], y=df[,3])) +
    scale_x_date(labels = date_format("%Y"), date_breaks='1 year') +
    geom_line(colour="blue", size=0.75) + 
    xlab('Year') + ylab(ylab) +
    theme(axis.text=element_text(size=10,face="bold",angle=90,hjust=1),
          axis.title=element_text(size=12,face="bold"),
          title=element_text(size=14,face="bold"))+
    ggtitle(title) 
  return(list(p1=p1,p2=p2))
}

compute_change <- function(df) {
  for(i in 1:(nrow(df)-1)) {
    y1 <- df[i,2]
    y2 <- df[i+1,2]
    change <- 100*(y2-y1)/y1
    df[i+1,3] <- change
  }
  return(df)
}

process <- function(df,period,index,xlabel,title) {

  ord <- order(df$time)
  df <- df[ord,]
  df <- df[complete.cases(df),]
  df <- df[,c('time','values')]
  start_time <- df$time[1]
  start_year <- year(start_time)
  start_month <- month(start_time)
  end_time <- df$time[nrow(df)]
  end_year <- year(end_time)
  end_month <- month(end_time)
  if(period=='q') { ## last is yyyy-10-01: extend to end of year
    all <- seq(start_time, as.Date(paste(end_year,'-12-01',sep='')), 
               by =  "month")
  } else if(period=='m') { ## last is yyyy-12-01: keep as is 
    all <- seq(start_time, end_time, by =  "month")
  } else if(period=='y') { ## last is yyyy-01-01: : extend to end of year
    all <- seq(start_time, as.Date(paste(end_year,'-12-01',sep='')), 
               by =  "month")
  }
  
  df2 <- data.frame(time=all)
  df2 <- merge(df,df2,all=TRUE)
  
  df2$values <- ts(df2$values,start=c(start_year,1),end=c(end_year,12),
                   frequency=12)
  df2$values <- na.interpolation(df2$values, option = "linear")

  df2 <- compute_change(df2)
  df2$V3 <- ts(df2$V3,start=c(start_year,1),end=c(end_year,12),frequency=12)

  names(df2)<-c('time',index,paste(index,'_change',sep=''))

  plots <- graph (df2,xlabel,title)
  print(plots$p1)
  print(plots$p2)

  return(list(df2=df2,p1=plots$p1,p2=plots$p2))
}

```


### Gross domestic product at market prices
### Price index (implicit deflator), 2010=100, Euro
### Seasonally and calendar adjusted data	
### Quarterly data
### 2000Q1 - 2016Q4


```{r, echo=TRUE}

gdp <- get_eurostat(id='namq_10_gdp',time_format = 'date',update_cache=TRUE,
                    filters= list(geo='BG',unit='PD10_EUR',s_adj='SCA',na_item='B1GQ'))

ll <- process(gdp,'q','GDP','Monthly Price index, 2010=100','GDP at market prices - Seasonally and calendar adjusted data') 
gdp2 <- ll$df2


```

### Nonperforming to total gross loans
### Yearly data
### 2000 - 2015
### World Bank Database
### http://data.worldbank.org/indicator/FB.AST.NPER.ZS?locations=BG


```{r, echo=TRUE}

npl <- WDI(country = c('BG'), indicator = "FB.AST.NPER.ZS",
start = 2000, end = 2015, extra = TRUE, cache = NULL)

npl <- npl[,c('year','FB.AST.NPER.ZS')] 
names(npl) <- c('time','values')
npl$time <- as.Date(paste(as.character(npl$time),'-01-01',sep=''))

ll <- process(npl,'y','NPL','% of gross loans','Non-performing loans')
npl2 <- ll$df2


dat <- merge(gdp2,npl2,by=c('time'),all = TRUE)
write.csv(dat,file='data.csv')

```




### Total unemployment rate: % of the active population 15-74 
### Quarterly data
### 2000Q1 - 2016Q4


```{r, echo=TRUE}


## Total unemployment rate: % of the active population
## LFS adjusted series

unempl <- get_eurostat(id='lfsq_urgan',time_format = 'date',update_cache=TRUE,
                    filters= list(age='Y15-74', citizen='TOTAL',sex='T', geo='BG'))


ll <- process(unempl,'q','Unempl','% of the active population','Total unemployment rate')
unempl2 <- ll$df2

dat <- merge(dat,unempl2,by=c('time'),all = TRUE)
write.csv(dat,file='data.csv')

```


### House price index (2010 = 100) - quarterly data
### Total purchases
### Quarterly data
### 2005Q1 - 2016Q4


```{r, echo=TRUE}
hpi <- get_eurostat(id='prc_hpi_q',time_format = 'date',update_cache=TRUE,
                    filters= list(unit='INX_Q', typpurch='TOTAL',geo='BG'))

ll <- process(hpi,'q','HPI','Index (2010 = 100)','House price index')
hpi2 <- ll$df2

dat <- merge(dat,hpi2,by=c('time'),all = TRUE)
write.csv(dat,file='data.csv')

```

### From the World Bank Database:
### Inflation, consumer prices (annual %)
### Yearly data
### 2000-2015
### http://data.worldbank.org/indicator/FP.CPI.TOTL.ZG?locations=BG


```{r, echo=TRUE}


inflat <- WDI(country = c('BG'), indicator = "FP.CPI.TOTL.ZG",
start = 2000, end = 2015, extra = TRUE, cache = NULL)
inflat <- inflat[,c('year','FP.CPI.TOTL.ZG')] ; 
names(inflat) <- c('time','values')

inflat$time <- as.Date(paste(as.character(inflat$time),'-01-01',sep=''))


ll <- process(inflat,'y','Inflation','Inflation','Inflation')
inflat2 <- ll$df2


dat <- merge(dat,inflat2,by=c('time'),all = TRUE)
write.csv(dat,file='data.csv')

```
## EMU convergence criterion bond yields - monthly data
## calculating Greek-German 10-year bond spread

```{r, echo=TRUE}
bonds <- get_eurostat(id='irt_lt_mcby_m',time_format = 'date',update_cache=TRUE)
bonds <- bonds[bonds$geo=='EL' | bonds$geo=='DE',]
bonds <- bonds[year(bonds$time) >= 2000,]
bonds <- dcast(bonds, time ~ geo, value.var = 'values')
bonds$values <- bonds$EL-bonds$DE
bonds <- bonds[,c('time','values')]

bonds$values <- ts(bonds$values,start=c(2000,1),end=c(2017,3),
                   frequency=12)

bonds$values <- na.interpolation(bonds$values, option = "linear")

bonds2 <- compute_change(bonds)
bonds2$V3 <- ts(bonds2$V3,start=c(2000,1),end=c(2017,3),frequency=12)

names(bonds2)<-c('time','Bonds','Bonds_change')

plots <- graph (bonds2,'bonds','bonds')
plots$p1
plots$p2
rm(bonds)

dat <- merge(dat,bonds2,by=c('time'),all = TRUE)
write.csv(dat,file='data.csv')

```


## average bank interest rates

##http://sdw.ecb.europa.eu/browseSelection.do?df=true&ec=&dc=&oc=&pb=&rc=&DATASE##T=0&removeItem=&activeTab=&REF_AREA=67&BS_ITEM=A20&BS_ITEM=A20&MATURITY_NOT_IR##ATE=&node=1513

```{r, echo=TRUE}
inter <- read.csv(file='interest_rates.csv',skip = 4,header=TRUE,
                  stringsAsFactors = FALSE)
names(inter) <- c('time','rate1','rate2','rate3')

inter$rate <- rowMeans(inter[,2:4])

inter$year <- as.numeric(substr(inter$time,1,4))
inter$month <- substr(inter$time,5,length(inter$time))

inter$time2 <- paste(inter$year,inter$month,'01',sep='-')
inter$time3 <- as.Date(inter$time2,format='%Y-%b-%d')

inter <- inter[, c('time3','rate')]
names(inter) <- c('time','values')

ord <- order(inter$time)
inter <- inter[ord,]

inter$values <- ts(inter$values,start=c(2007,1),end=c(2017,2),
                   frequency=12)

inter2 <- compute_change(inter)
inter2$V3 <- ts(inter2$V3,start=c(2007,1),end=c(2017,2),frequency=12)

names(inter2)<-c('time','Inter','Inter_change')

plots <- graph (inter2,'inter2','inter2')
plots$p1
plots$p2
rm(inter)

dat <- merge(dat,inter2,by=c('time'),all = TRUE)
write.csv(dat,file='data.csv')


```
```{r, echo=TRUE}
for(j in 2:ncol(dat) )
  dat[,j] <- ts(dat[,j], start=c(2000,1), end=c(2017,3), frequency = 12)

model0 <- lm(NPL ~ . -time -NPL_change, data=dat)
summary(model0)


```
model0 <- tslm(non_perf_loans ~ ., data=dat)




summary(model0)
vif(model0)
dat8$preds0 <- predict(model0,data=dat8)

autoplot(cbind(Actual=dat8$non_perf_loans, Predicted=dat8$preds0), ts.linetype = 'solid',ylab='% Non-performing loans',ts.size=1)



ggplot(dat8, aes(x = non_perf_loans, y = preds0)) + 
    #geom_line(colour="blue", size=0.75) + 
    scale_x_continuous(breaks=seq(5,25,5))+
    scale_y_continuous(breaks=seq(5,25,5))+
    geom_point()  +
    geom_abline(intercept = 0, slope = 1, color='blue')+
    xlab('Non-performing loans ratio (actual)') + 
    ylab('Non-performing loans ratio (predicted)') 
    #facet_wrap(~ variable, scales = 'free_y', ncol = 1)

plot(model0)

```

We then tested each predictor separately. The p-values showed significant associations of the unemployment rate, inflation rate and the Greek-German bonds spread.
```{r, eval=FALSE, echo=FALSE}

for( j in 2:7) {
  cat('\npredictor: ',names(dat8)[j],'\n')
  model <- tslm(non_perf_loans ~ dat8[,j], data=dat8)
  print(summary(model)$coefficients[2,4])
  stat <- cor.test(dat8$non_perf_loans,dat8[,j])
  print(stat$p.value)
}
```

After some tests, we also excluded the unemployment rate and concluded with a model with two predictors, inflation rate, the Greek-German bonds spread and the average bank interest rate. This model has a quite good fit (adjusted R2 = 0.84) and diagnostics, as presented below.  

```{r, eval=FALSE, echo=FALSE}

model1 <- tslm(non_perf_loans ~ Inflation + int_rate
                 + spread, data=dat8)
summary(model1)
vif(model1)
dat8$preds1 <- predict(model1,data=dat8)
spreadLevelPlot(model)

plot(model1)


ggplot(dat8, aes(x = non_perf_loans, y = preds1)) + 
    #geom_line(colour="blue", size=0.75) + 
    scale_x_continuous(breaks=seq(5,25,5))+
    scale_y_continuous(breaks=seq(5,25,5))+
    geom_point()  +
    geom_abline(intercept = 0, slope = 1, color='blue')+
    xlab('Non-performing loans ratio (actual)') + 
    ylab('Non-performing loans ratio (predicted)') 
    #facet_wrap(~ variable, scales = 'free_y', ncol = 1)

autoplot(cbind(Actual=dat8$non_perf_loans, Predicted=dat8$preds1), ts.linetype = 'solid',ylab='% Non-performing loans',ts.size=1)


```

## References
***

Jakubík, Petr, and Thomas Reininger. What are the key determinants of nonperforming loans in CESEE?. No. 26/2014. IES Working Paper, 2014. 

Vogiazas, Sofoklis D., and Eftychia Nikolaidou. "Investigating the determinants of nonperforming loans in the Romanian banking system: An empirical study with reference to the Greek crisis." Economics Research International 2011 (2011).

CIA/ The World FactBook  
https://www.cia.gov/library/publications/the-world-factbook/geos/bu.html

## TO CHECK:
*** 
https://www.eba.europa.eu/documents/10180/1018121/Jakubik,%20Reininger+-+What+are+the+Key+Determinants+of+Nonperforming+Loans+in+CESEE+-+Paper.pdf

https://www.hindawi.com/journals/ecri/2011/214689/

http://ec.europa.eu/europe2020/pdf/csr2015/cr2015_bulgaria_en.pdf